Big Ideas in Sports Analytics and Statistical Tools for their Investigation

### 2022-11-28 sports_analytics.Rmd
close #29

Most of the research in sports analytics is fueled by the analysis of data recorded from the outcome of sports contests.
However, a growing body of literature is informed by non-performance data about sports. 
This research includes fields like sports economics and kinesiology that are subfields of mature disciplines (a treatment of either is well beyond the scope of the this paper).
In addition to this body of research, we note that today, professional sports teams are collecting biometric data on their own players, and presumably using that data to learn about how their workouts, sleep patterns, and diets impact their athletic performance. 
While this research may constitute "emerging methodologies," it will likely never be published, because each team will use it to their competitive advantage.

Nevertheless, in this Section we discuss concepts stemming from one notable, relatively new source of (semi-)public non-performance sports data: data from betting markets.
Since the 2018 United States Supreme Court decision in [*Murphy v. National Collegiate Athletic Association*](https://en.wikipedia.org/wiki/Murphy_v._National_Collegiate_Athletic_Association), sports gambling has exploded in the U.S. 
The increasing interest in sports gambling has led to increasing interest in sports gambling *data*, and that data has proven useful to researchers in two major ways.

@@

Second, sports analytics researchers have studied various types of sports gambling outlets (including fantasy sports). 
This research has estimated probabilities, evaluated common strategies, and offered optimized approaches for a variety of different games of chance (see Section \ref{sec:pools}).

## Sports betting market data {#sec:betting-market-data}

While @Lopez2018 uses betting market data to model team strengths, it does not directly address strategies for betting or inefficiencies in betting markets. 
Work such as @lacey1990estimation, @Boulier2006,

@Boulier2006 that there is insufficient evidence to conclude that betting markets were inefficient from 1994--2000.
@boulier2003predicting test several predictors for NFL games and find that the betting market is the best predictor.
@Buttrey2016 demonstrate a method that produces a positive ROI.
@gandar1988testing conclude that stats test are too weak to reject the hypothesis of rationality.  But they do reject rationalist with ``economic tests". 
@lacey1990estimation : Finds the market to be efficient, though they did find some profitable betting rules over the period of time between 1984 and 1986

## Analytics for sports betting pools {#sec:pools}

Another popular type of betting pool is a survivor pool.
@Bergman2017SurvivingAN presents formal optimization approaches for NFL survivor pools and concludes that planning for only part of the season yields optimal results in terms of maximizing survival probability.

@breiter1997play

@clair2007optimal

@kaplan2001march

@metrick1996march

@niemi2008contrarian

@sauer1998economics

@imbrogno2022computing
Description: The average number of winning survivor pool entries is computed for the 2019 National Football League (NFL) season. The probability of having to share the winning pot, given realistic expectations on game outcomes, is also shown.

@nichols2014impact conclude that the impact of travel is not completely incorporated into the betting markets, but that any effect is too small to find any profitable advantage.

@PaulWeinbach2014 Market efficiency and behavioral biases in the wnba betting market

@soebbing2013gamblers

@spann2009sports NBA bettors believe that tanking is occurring.

>>>>

Most of the research in sports analytics is fueled by the analysis of data recorded from the outcome of sports contests.
However, a growing body of literature is informed by non-performance data about sports. 
This research appears in fields like sports economics and kinesiology that are subfields of mature disciplines (a treatment of either is well beyond the scope of the this paper).
In addition to this body of research, we note that today, professional sports teams are collecting biometric data on their own players, and presumably using that data to learn about how their workouts, sleep patterns, and diets impact their athletic performance. 
While this research may constitute "emerging methodologies," it will likely never be published, because each team will use it to their competitive advantage.

Nevertheless, in this Section we discuss concepts stemming from one notable, relatively new source of (semi-)public non-performance sports data: betting markets.
Since the 2018 United States Supreme Court decision in [*Murphy v. National Collegiate Athletic Association*](https://en.wikipedia.org/wiki/Murphy_v._National_Collegiate_Athletic_Association), sports gambling has exploded in the U.S. 
The increasing interest in sports gambling has led to increasing interest in sports gambling *data*, and that data has proven useful to researchers in at least two major ways.

@@

Second, sports analytics researchers have studied various types of sports gambling outlets (including fantasy sports). 
This research has estimated probabilities, evaluated common strategies, and offered optimized approaches for a variety of different games of chance (see Section \ref{sec:pools}). 
Some researchers have then tried to demonstrate a positive return on some of these betting strategies, with very limited success.

## The use of sports betting market data {#sec:betting-market-data}

While @Lopez2018 uses betting market data to model team strengths, they do not directly address strategies for betting or inefficiencies in betting markets. 
Early work by @gandar1988testing examines the rationality of NFL betting markets and concludes that statistical tests fail to reject the hypothesis of rationality.
Related work such as @lacey1990estimation and @Boulier2006 explores the efficiency of NFL betting markets in the mid-1980s and late-1990s, respectively.
Neither paper finds strong evidence for inefficiencies in the markets. 
@boulier2003predicting compare the predictive performance of power rankings and media experts to the betting market for NFL games and found that the betting market is the best for predicting winners. 
@lopez2015building show that betting market data was most useful in predicting men's college basketball tournament outcomes.

Sports betting market data has also been used to investigate competitive behavior within leagues. 
@soebbing2013gamblers find evidence that sports bettors *think* tanking in the NBA is occurring, although the evidence for whether it actually is remains mixed.

## Analytics for sports betting {#sec:pools}

### Betting pools

One popular type of betting pool is a survivor pool, in which participants stay in the competition as long as they continue to successfully pick winners.
@Bergman2017SurvivingAN present formal optimization approaches for NFL survivor pools and conclude that planning for only part of the season yields optimal results in terms of maximizing survival probability.
@imbrogno2022computing estimate the probability of having to share the winning pot in NFL survivor pools.

Perhaps the most commonly-studied sports betting market surrounds the NCAA men's college basketball tournament. 
@breiter1997play use Monte Carlo methods to study the standard "office pool."
@kaplan2001march consider a variety of NCAA college basketball pools, and find that the simple strategy of picking the team with the better seed is generally, but not always, optimal. 
@metrick1996march finds that bettors overback the heaviest favorites. 
@niemi2008contrarian show an improved return on investment by picking an undervalued champion and then completing the rest of one's bracket by using published odds. 
@clair2007optimal develop and test strategies for maximizing expected return in both survivor and tournament-style pools.

### Beating the market

Naturally, after studying the efficiency of sports betting markets, researchers try to find inefficiencies that can be exploited for financial gain. 
Not surprisingly (given the efficiency of these markets), such gains are hard to come by.

@sauer1998economics finds that while racetrack betting markets are generally efficient, information asymmetry plays a role in creating inefficient markets. 
@nichols2014impact concludes that the impact of travel is not completely incorporated into the betting markets, but that any effect is too small to find any profitable advantage. 
@PaulWeinbach2014 investigate the less-saturated betting market for the WNBA and fail to find strategies for positive return on investment. 
@spann2009sports show no way to beat the market in the German premier soccer league, given the high fees associated with placing bets.

More successfully, @Buttrey2016 explores the NHL betting market and produces a model to predict win probabilities in given games, then tests the model by placing market price bets in games where the predicted probability differs from the market. 
They find that their methods were able to produce a positive return on investment.
